---
title: "What are boosted regression trees?"
author: "Andrew M. Dolman"
date: '`r format(Sys.time(), "%d %B, %Y")`'
output:
  ioslides_presentation:
    css: styles.css
institute: Brandenburg University of Technology, Cottbus - Senftenberg
---

```{r knitr_opts, echo=FALSE, warning=FALSE, message=FALSE, echo=FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE, tidy=FALSE, echo = FALSE, cache=TRUE)
```

```{r packages}
library(gbm)
library(ggplot2)
library(dplyr)
library(tidyr)
library(gridExtra)
library(tree)
library(ggdendro)
```

```{r functions}
gg_arrange <-
  function(lst, grid.arrange_args = NULL, labels = NULL, align_horizontal = TRUE, align_vertical = TRUE, single_y_axis = FALSE) {
    # Add labels
    if (is.null(labels) == FALSE) {
      fun <- function(lst, labels) {
        lst <-
          lst + labs(title = labels) +  theme(plot.title = element_text(hjust = 0, vjust = 1))
        return(lst)
      }
      
      lst <- mapply(fun, labels = labels, lst = lst, SIMPLIFY = FALSE)
    }
    
    # Convert to gtables
    tables <- lapply(lst, function(x)
      ggplot2::ggplot_gtable(ggplot2::ggplot_build(x)))
    
    if (align_horizontal == TRUE) {
      # Get max width and set widths of all plots to this
      widths <- lapply(tables, function(x)
        x$widths)
      maxWidth <- do.call(grid::unit.pmax, widths)
      tables <- lapply(tables, function(x) {
        x$widths <- maxWidth
        x
      })
    }
    
    if(single_y_axis == TRUE){
      if(length(lst) != 2) stop("binding with single y axis only supported for 2 column figure at present")
      
      # get width of first 2 cols
      sum_lft <- sum(grid::convertWidth(tables[[2]]$widths[1:2], "cm", valueOnly = T))
      
      # create new out boundary
      tables[[2]]$widths[7] <- tables[[2]]$widths[6]
      
      # replace previous boundary with width of 1:2
      tables[[1]]$widths[6] <- grid::unit(sum_lft, "cm")
      
      grid::convertWidth(tables[[2]]$widths[6], "cm")
      # set first 2 column to 0 width (axis label and numbers)
      tables[[2]]$widths[1:2] <- 0  
      }
    
    if (align_vertical == TRUE) {
      # Do the same for heights
      heights <- lapply(tables, function(x)
        x$heights)
      maxHeight <- do.call(grid::unit.pmax, heights)
      tables <- lapply(tables, function(x) {
        x$heights <- maxHeight
        x
      })
    }
    
    # Build the final plot
    args_list <- c(tables, grid.arrange_args)
    g <- do.call(gridExtra::arrangeGrob, args_list)
    return(g)
  }

theme_bw_2 <- function() {ggplot2::theme_bw() +
    ggplot2::theme(panel.grid.minor = ggplot2::element_blank())}

theme_bw_2_asp1 <- function() {ggplot2::theme_bw() +
    ggplot2::theme(panel.grid.minor = ggplot2::element_blank(), aspect.ratio = 1)}

```


```{r make_data}
make_data <- function(n, Grp = "A", b0 = 5, b1 = -0.1){

x = seq(0.01, 10, length.out = n)
dat <- data.frame(x = x,
                  Grp = Grp,
                  True = sin(x) + b0 + b1 * x,
                  y = sin(x) + rnorm(length(x), 0, 0.5) + b0 + b1 * x)


# Define test and training sets

dat$Set <- FALSE
dat[sample(1:nrow(dat), nrow(dat) / 5, replace = FALSE), "Set"] <- "Test"
dat$Set <- factor(ifelse(dat$Set == "Test", "Test", "Training"), ordered = TRUE,
                  levels = c("Training", "Test"))  
dat <- dat[(order(dat$Set)),]
return(dat)
}

set.seed(1)
dat <- make_data(n = 1000) %>% 
  tbl_df()

#dat$x2 <- factor("A")
```


```{r fit_gbms, results='hide', message = FALSE, warning=FALSE, echo = FALSE, error = FALSE}
shrinks <- data.frame(shrinkage = c(1, 0.5, 0.1, 0.01))

mods <- plyr::alply(shrinks, 1, function(s) gbm(y ~ x, shrinkage = s$shrinkage
                                                   , n.trees = 5000
                                                   , train.fraction = 4/5
                                                   , n.minobsinnode = 5
                                                   , data = dat, verbose = FALSE))

par(mfrow = c(2,2))
best_iter <- plyr::ldply(mods, gbm.perf, method = "test", plot.it = FALSE)
par(mfrow = c(1, 1))
```


```{r newdata}
newdat <- dat %>% 
  arrange(x)

preds <- plyr::ldply(mods, function(x) predict(x, newdata = newdat, n.trees = c(1:8, 16, 32, 64, 128, 1000, 5000)))

newdat2 <- cbind(newdat, preds) %>% 
  gather(ntrees, prediction, -x, -Grp, -y, -Set, -True, -shrinkage) %>%
  mutate(resid = y - prediction,
         tree_grp = "seq",
         ntrees = as.integer(as.character(ntrees))) %>% 
  tbl_df()


best_preds <- plyr::ldply(mods, function(x)
  data.frame(prediction = predict(x, newdata = newdat, n.trees = gbm.perf(x, method = "test", plot.it = FALSE)),
             ntrees = (gbm.perf(x, method = "test", plot.it = FALSE)),
             tree_grp = "best")) %>% 
  cbind(newdat) %>% 
  tbl_df()


newdat3 <- bind_rows(newdat2, best_preds) %>% 
  mutate(shrink_num = shrinkage, 
         shrinkage = factor(shrinkage, ordered = T, levels = (unique(shrinkage))),
         ntrees = factor(ntrees, ordered  = T)) 

newdat_all_x <- newdat3 %>% 
  filter(Set == "Training") %>% 
  mutate(Set = "Test") %>% 
  bind_rows(., filter(newdat3, Set == "Training")) %>% 
  select(Set, x, y, True, prediction, shrinkage, shrink_num, ntrees)

## Progressive fitting

newdat4 <- newdat3 %>% 
  filter(ntrees == 1) %>% 
  mutate(ntrees = factor(0),
         prediction = mean(y),
         resid = y - prediction) %>% 
  bind_rows(newdat3, .)

newdat5 <- newdat4 %>%
  filter(ntrees %in% 0:8) %>%
  select(-resid) %>% 
  spread(ntrees, prediction) %>% 
  mutate(`0` = `1` - `0`,
         `1` = `2` - `1`,
         `2` = `3` - `2`,
         `3` = `4` - `3`,
         `4` = `5` - `4`,
         `5` = `6` - `5`,
         `6` = `7` - `6`,
         `7` = `8` - `7`
  ) %>% 
  gather(ntrees, next_tree, -tree_grp, -shrinkage, -shrink_num, -Grp, -Set, -y, -True, -x)

newdat6 <- newdat4 %>% 
#select(x, resid, ntrees, shrinkage, Set) %>% 
  left_join(., newdat5) %>% 
  mutate(tree = as.numeric(ntrees) + 1)
```

# An example

## Estimate unknown function

Simulate some data

```{r show_sim_data}
newdat %>%
  ggplot(aes(x = x, y = y)) %>% 
  + geom_line(aes(y = True), colour = "Black", size = 1.5) %>% 
  + geom_point(colour = "Black", alpha = 0.5) %>% 
#  + facet_wrap(~ Set, labeller = label_both)%>% 
  + theme_bw_2_asp1()
```



```{r basic_tree, include=FALSE}
tr1 <- tree(y ~ x, data = newdat)

# tr2 <- tr1
# tr2$frame$splits[,1] <- ""
plot(tr1, type = "uniform")
text(tr1, label = "yval", digits = 2, all = T)
#text(tr1, label = "dev", digits = 2, all = T)

tree_data <- dendro_data(tr1, type = "proportional")
tree_segs <- segment(tree_data)

p_tree <- tree_segs %>%
  ggplot() %>%
  + geom_segment(aes(x = x, y = y, xend = xend, yend = yend)) %>%
#  + geom_text(data = label(tree_data), aes(x = x, y = y, label = label), vjust = -1) %>%
  + geom_text(data = leaf_label(tree_data), aes(x = x, y = y, label = round(label, 1)), vjust = 1) %>%
  + theme_dendro() %>%
  + scale_x_continuous(limits = c(0, 10)) %>%
  + theme(aspect.ratio = 0.75)
```

## 

With one predictor a tree can be represented as a piecewise function

```{r}
newdat$y_hat_tr1 <- predict(tr1)

p <- newdat %>%
  ggplot(aes(x = x, y = y)) %>% 
  + geom_point(colour = "Black", alpha = 0.25) %>% 
  + geom_line(aes(y = True, colour = "True function"), size = 1.5) %>% 
  + geom_line(aes(y = y_hat_tr1, colour = "Tree estimate"), size = 1.5) %>% 
  + scale_colour_manual("", values = c("Red", "Black")) %>% 
  + scale_x_continuous(limits = c(0, 10)) %>% 
  + theme_bw() %>% 
  + theme(aspect.ratio = 1) %>% 
  + theme(legend.position = "top")
#p

p12 <- arrangeGrob(p_tree, p, ncol = 2)
grid::grid.newpage()
grid::grid.draw(p12)
```



```{r tree_1}
tree_seq <- function(fun_dat, tree_dat, trees, shrink){
  p1 <- fun_dat %>%
  filter(shrinkage %in% as.numeric(shrink), 
         Set == "Test", 
         ntrees %in% c(trees)) %>% 
      droplevels(.) %>% 
  ggplot(aes(x = x, y = y)) %>% 
  + geom_point() %>% 
  + geom_line(aes(y = prediction), colour = "#F21A00", size = 1.5) %>% 
  + facet_grid(ntrees ~ shrinkage, labeller=label_both) %>% 
  + theme_bw() %>%  
  + theme(aspect.ratio = 1) %>% 
  + theme(aspect.ratio = 1) %>% 
  + labs(title = "Sum of trees")


p2 <- tree_dat %>%
  filter(shrinkage %in% as.numeric(shrink), 
         Set == "Test", 
         ntrees %in% c(trees)) %>% 
  droplevels(.) %>% 
  ggplot(aes(x = x, y = resid)) %>% 
  + geom_point() %>% 
  + geom_line(aes(y = next_tree / as.numeric(shrink_num)), colour = "#3B9AB2", size = 1.5) %>% 
  + facet_grid(tree ~ shrinkage, labeller=label_both) %>% 
  + theme_bw() %>% 
  + theme(aspect.ratio = 1) %>% 
  + scale_y_continuous("Residuals") %>% 
  + labs(title = "Next tree")

p12 <- limnolrgy::gg_arrange(list(p1, p2), grid.arrange_args = c(ncol = 2), align_horizontal = TRUE, align_vertical = TRUE)

return(p12)
}
```


## Zero trees - (single split)

```{r}
p12 <- tree_seq(newdat4, newdat6, trees = 0, shrink = 1)
grid::grid.newpage()
grid::grid.draw(p12)
```



## 1 tree 

```{r tree_2}
p12 <- tree_seq(newdat4, newdat6, trees = 1, shrink = 1)
grid::grid.newpage()
grid::grid.draw(p12)
```

## 2 trees 

```{r tree_3}
p12 <- tree_seq(newdat4, newdat6, trees = 2, shrink = 1)
grid::grid.newpage()
grid::grid.draw(p12)
```

## 3 trees 

```{r tree_4}
p12 <- tree_seq(newdat4, newdat6, trees = 3, shrink = 1)
grid::grid.newpage()
grid::grid.draw(p12)
```

## 4 trees 

```{r tree_5}
p12 <- tree_seq(newdat4, newdat6, trees = 4, shrink = 1)
grid::grid.newpage()
grid::grid.draw(p12)
```

## 5 trees

```{r tree_6}
p12 <- tree_seq(newdat4, newdat6, trees = 5, shrink = 1)
grid::grid.newpage()
grid::grid.draw(p12)
```

## 6 trees

```{r tree_7}
p12 <- tree_seq(newdat4, newdat6, trees = 6, shrink = 1)
grid::grid.newpage()
grid::grid.draw(p12)
```

## 7 trees

```{r tree_8}
p12 <- tree_seq(newdat4, newdat6, trees = 7, shrink = 1)
grid::grid.newpage()
grid::grid.draw(p12)
```



## 0 - 3 trees 

```{r trees_1_4, fig.width=5, fig.height=5.5}
p12 <- tree_seq(newdat4, newdat6, trees = 0:3, shrink = 1)
grid::grid.newpage()
grid::grid.draw(p12) 
```

## > Lots of trees

```{r oob, fig.width=5}
newdat3 %>%
  filter(ntrees %in% c(1, 8, 64, 1000),
         shrinkage == 1,
         Set == "Test") %>% 
  ggplot(aes(x = x, y = prediction)) %>% 
  + geom_point(aes(x=x, y=y), colour = "Darkgrey") %>% 
  + geom_line(aes(y = True), colour = "Black", size = 1.2) %>% 
  + geom_line(colour = "#F21A00", size = 1.2) %>% 
  + facet_wrap(~ntrees, ncol = 2, labeller=label_both) %>% 
  + theme_bw_2()
```

# When to stop?

## Use test data

```{r oob_over, fig.width=7}
newdat_all_x_sub <- newdat_all_x %>% 
  filter(ntrees %in% c(5000),
         shrinkage == 1,
         x  < 0.5, x > 0.1) %>% 
  mutate(Set = factor(Set, ordered = T, levels = c("Training", "Test")))


newdat3 %>%
  filter(ntrees %in% c(5000),
         shrinkage == 1,
         x  < 0.5, x > 0.1) %>% 
  ggplot(aes(x = x, y = prediction)) %>% 
  + geom_point(aes(x=x, y=y), colour = "Black") %>% 
  + geom_line(data = newdat_all_x_sub, aes(y = True), colour = "Black") %>% 
  + geom_line(data = newdat_all_x_sub, colour = "#F21A00") %>% 
  + facet_grid(ntrees~Set, labeller=label_both) %>% 
  + theme_bw_2()
```



```{r}
get_error <- function(mod) {dat <- with(mod, data.frame(n_trees = 1:n.trees, train_error = train.error, test_error = valid.error)) %>%
  gather(error_type, error, -n_trees)
return(dat)
}

error_df <- plyr::ldply(mods, get_error) %>% 
  tbl_df()
```

## Test vs. training error

```{r error_shrink_1}
error_df %>% 
  filter(n_trees <= 500,
         shrinkage == 1) %>% 
  ggplot(aes(x = n_trees, y = error, colour = error_type)) %>% 
  + geom_line() %>% 
  + theme_bw_2()
```

## Optimum no. trees

Still wiggly

```{r opt_shrink_1}
newdat3 %>%
  filter(tree_grp == "best",
         shrinkage == 1,
         Set == "Test") %>% 
  ggplot(aes(x = x, y = prediction)) %>% 
  + geom_point(aes(x=x, y=y)) %>% 
  + geom_line(colour = "#F21A00", size = 1.5) %>% 
  + facet_wrap(~ shrinkage + ntrees, labeller=label_both) %>% 
  + theme_bw_2()
```


# Lower the learning rate (shrinkage)

## Zero trees, LR = 0.1

```{r tree_0_0.5}
p12 <- tree_seq(newdat4, newdat6, trees = 0, shrink = 0.1)
grid::grid.newpage()
grid::grid.draw(p12)
```

## 1 tree, LR = 0.1

```{r tree_2_0.5}
p12 <- tree_seq(newdat4, newdat6, trees = 1, shrink = 0.1)
grid::grid.newpage()
grid::grid.draw(p12)
```

## 2 trees, LR = 0.1

```{r tree_3_0.5}
p12 <- tree_seq(newdat4, newdat6, trees = 2, shrink = 0.1)
grid::grid.newpage()
grid::grid.draw(p12)
```

## 3 trees, LR = 0.1

```{r tree_4_0.5}
p12 <- tree_seq(newdat4, newdat6, trees = 3, shrink = 0.1)
grid::grid.newpage()
grid::grid.draw(p12)
```

## 4 trees, LR = 0.1

```{r tree_5_0.5}
p12 <- tree_seq(newdat4, newdat6, trees = 4, shrink = 0.1)
grid::grid.newpage()
grid::grid.draw(p12)
```

## 5 trees, LR = 0.1

```{r tree_6_0.5}
p12 <- tree_seq(newdat4, newdat6, trees = 5, shrink = 0.1)
grid::grid.newpage()
grid::grid.draw(p12)
```

## 6 trees, LR = 0.1

```{r tree_7_0.5}
p12 <- tree_seq(newdat4, newdat6, trees = 6, shrink = 0.1)
grid::grid.newpage()
grid::grid.draw(p12)
```

## 7 trees, LR = 0.1

```{r tree_8_0.5}
p12 <- tree_seq(newdat4, newdat6, trees = 7, shrink = 0.1)
grid::grid.newpage()
grid::grid.draw(p12)
```


## 0 - 3 trees 

```{r trees_1_4_0.1, fig.width=5, fig.height=5.5}
p12 <- tree_seq(newdat4, newdat6, trees = c(0, 1, 2), shrink = 0.1)
grid::grid.newpage()
grid::grid.draw(p12) 
```


## Test error

```{r error_shrink_2}
error_df %>%
  mutate(shrinkage = factor(shrinkage, ordered = T, levels = (unique(shrinkage)))) %>% 
  filter(n_trees <= 1000,
         shrinkage %in% c(0.1, 1)) %>% 
  ggplot(aes(x = n_trees, y = error, colour = error_type)) %>% 
  + geom_line() %>% 
  + facet_wrap(~shrinkage, as.table = TRUE, labeller=label_both) %>% 
  + theme_bw_2() %>% 
  + geom_hline(yintercept = filter(error_df, shrinkage=="0.1"
                                   , n_trees == best_iter[best_iter$shrinkage=="0.1", "V1"]
                                   , error_type == "test_error")$error) %>% 
  + scale_y_continuous("Test error")
```


## Optimum trees; LR = 1 vs. 0.1 
```{r opt_shrink_0.1}
newdat3 %>%
  filter(tree_grp == "best",
         shrinkage %in% c(0.1, 1),
         Set == "Test") %>% 
  ggplot(aes(x = x, y = prediction)) %>% 
  + geom_point(aes(x=x, y=y), colour = "Darkgrey") %>% 
  + geom_line(aes(y = True), colour = "Black", size = 1.5) %>%
  + geom_line(colour = "#F21A00", size = 1.5) %>%
  + facet_wrap(~ shrinkage + ntrees, labeller=label_both) %>% 
  + theme_bw_2()
```


# Interaction depth

```{r make_grp_data}
tmp <- make_data(n = 1000, Grp = "B", b0 = 4, b1 = 0.025) %>% 
  tbl_df()

dat2 <- bind_rows(dat, tmp) %>% 
  mutate(Grp = factor(Grp))
```

##

```{r}
p <- dat2 %>%
  ggplot(aes(x = x, y = y, colour = Grp)) %>% 
  + geom_line(aes(y = True), size = 1.5) %>% 
  + geom_point(alpha = 0.5) %>% 
  + theme_bw_2_asp1()
p
```

# Fitting a BRT with gbm()

## Data

```{r look_at_data}
dat2 %>% 
  select(-Set, -True) %>% 
  arrange(x, Grp)
```


## Two groups (interaction.depth = 1)

interaction.depth = 1 means ***no interaction!***

```{r brt_id_1, echo = TRUE}
brt_id_1 <- gbm (y ~ x + Grp, data = dat2
            , n.trees = 1000
            , shrinkage = 0.1
            , interaction.depth = 1
            , cv.folds = 5)
```

## Model diagnostics
```{r, echo=T, tidy=TRUE, tidy.opts=list(blank=TRUE, width.cutoff=40)}
brt_id_1
```

## 
```{r, echo=T, fig.show='hold', fig.height=3.5, fig.width=4.5}
best_iter <- gbm.perf(brt_id_1)
print(best_iter)
```

##
```{r, echo=T, fig.width=3, fig.show='hold', fig.height=3}
summary(brt_id_1, n.trees = best_iter)
```

##
```{r, echo=T}
plot(brt_id_1
     , n.trees = best_iter
     , i.var = 1:2, layout = c(2, 1))
```



## , interaction.depth = 1
```{r, echo = TRUE}
dat2$y_hat_id_1 <- predict(brt_id_1, n.trees = best_iter)
```

```{r}
p <- dat2 %>%
  ggplot(aes(x = x, y = y, colour = Grp)) %>% 
  + geom_point(alpha = 0.25) %>% 
  + theme_bw_2_asp1() %>% 
  + theme(legend.position = "none")

p1 <- p + 
    geom_line(aes(y = True, colour = Grp), size = 1.5) +
    ggtitle("True function")

p2 <- p +
    geom_line(aes(x = x, y = y_hat_id_1, colour = Grp), size = 1.5) +
    ggtitle("BRT estimate")

p12 <- gridExtra::grid.arrange(p1, p2, ncol = 2)
```


## Two groups (interaction.depth = 2)


```{r brt_id_2, echo = TRUE}
brt_id_2 <- gbm(y ~ x + Grp, data = dat2
            , n.trees = 1000
            , shrinkage = 0.1
            , interaction.depth = 2
            , cv.folds = 5)

best_iter2 <- gbm.perf(brt_id_2, plot.it = F)
```


## , interaction.depth = 2
```{r, echo = TRUE}
dat2$y_hat_id_2 <- predict(brt_id_2, n.trees = best_iter2)
```

```{r}
p <- dat2 %>%
  ggplot(aes(x = x, y = y, colour = Grp)) %>% 
  + geom_point(alpha = 0.25) %>% 
  + theme_bw_2_asp1() %>% 
  + theme(legend.position = "none")

p1 <- p + 
    geom_line(aes(y = True, colour = Grp), size = 1.5) +
    ggtitle("True function")

p2 <- p +
    geom_line(aes(x = x, y = y_hat_id_2, colour = Grp), size = 1.5) +
    ggtitle("BRT estimate")

p12 <- gridExtra::grid.arrange(p1, p2, ncol = 2)
```

## Summary boosted trees

* Sum of long chain of simple trees iteratively fit to residuals
    * n.trees = how many trees to fit
    * shrinkage = learning rate ≊ regularization
    * interaction.depth ≊ tree depth ≊ base learner complexity
* Optimal size and complexity from test error (out of bag, OOB)
    * true test set or cross-validation
* R package *gbm* shown here - others available

## Resources

Free pdf versions of these great text books are available from the authors.  

*  "The Elements of Statistical Learning: Data Mining, Inference, and Prediction" Hastie, T., Tibshirani, R., & Friedman, J. 
    * http://statweb.stanford.edu/~tibs/ElemStatLearn/


*  "An Introduction to Statistical Learning with Applications in R" Gareth James, Daniela Witten, Trevor Hastie, Robert Tibshirani 
    * http://www-bcf.usc.edu/~gareth/ISL/


